---
title: "Introduction to Linkmapper"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Linkmapper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Overview

Linkmapper is a free, open-source Shiny web application for linkage mapping and
QTL visualisation in biparental mapping populations. It wraps the
[onemap](https://cran.r-project.org/package=onemap) R package and the
[qtl](https://cran.r-project.org/package=qtl) package behind a graphical
interface, allowing users to move from raw genotype data to a finalised linkage
map and QTL scan without writing any R code.

The application is structured as a five-step guided workflow. Each step unlocks
only after the previous step has been completed successfully, preventing out-of-order
analysis.

```{r workflow, echo=FALSE, message=FALSE}
if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz("
    digraph workflow {
      graph [rankdir=LR, fontname=Helvetica]
      node [shape=box, style=filled,
            fontname=Helvetica, fontsize=11]

      A [label='Upload .txt', fillcolor='#fef3e2']
      B [label='Prior analysis', fillcolor='#e8f2ec']
      C [label='Marker grouping', fillcolor='#e8f2ec']
      D [label='Ordering', fillcolor='#e8f2ec']
      E [label='Linkage map', fillcolor='#e8f2ec']
      F [label='QTL analysis', fillcolor='#e8f2ec']

      A->B->C->D->E->F
    }
  ")
} else {
  message("Install DiagrammeR to render the workflow diagram: install.packages('DiagrammeR')")
}
```

---

## Installation

### Hosted web app

The easiest way to use Linkmapper is through the hosted version on ShinyApps.io.
No installation is required — open the URL in any modern browser and begin uploading
your data immediately.

<!-- FILL: add ShinyApps.io URL -->

### R package (local)

To run Linkmapper locally, install it from R-universe:

```r
install.packages("linkmapper",
  repos = "https://<!-- FILL: username -->.r-universe.dev"
)

linkmapper::run_linkmapper()
```

This opens the app in your default browser. Running locally removes the session
time limits of the hosted version and allows you to work with larger datasets.

---

## Data requirements

### File format

Linkmapper accepts genotype data in **MAPMAKER `.txt` format**. The file must begin
with a header line specifying the data type, followed by a line declaring the number
of individuals, markers, and phenotype columns:

```
data type f2 intercross
188 62 2
```

Supported data type strings:

- `f2 intercross` — F2 intercross population
- `backcross` — backcross population

### Genotype encoding

Genotypes are encoded per MAPMAKER conventions:

| Code | F2 meaning | Backcross meaning |
|------|------------|-------------------|
| `A`  | Homozygous parent 1 | Parent 1 homozygote |
| `B`  | Homozygous parent 2 | Backcross heterozygote |
| `H`  | Heterozygous | (not used) |
| `-`  | Missing | Missing |

### Phenotype columns

Phenotype columns are optional for linkage mapping but are **required for QTL
analysis**. If your file contains phenotype data, it must be declared in the header
(the third number on the second line). Phenotype values should be numeric; missing
phenotype data should be encoded as `-`.

---

## The five-step workflow

### Step 1 — Upload and prior analysis

Upload your MAPMAKER `.txt` file. Linkmapper validates the file format and reads the
data using `onemap::read_mapmaker()`. The prior analysis panel then displays:

- A **missing data plot** showing the proportion of missing genotype calls per marker
  and per individual, helping identify low-quality markers or individuals to consider
  removing before proceeding.
- A **segregation distortion test** (chi-squared) for each marker. Markers that
  deviate significantly from expected Mendelian ratios are flagged. Users can choose
  whether to retain or exclude distorted markers before grouping.

At the end of this step, the number of individuals, markers, and phenotype columns
is summarised, and a suggested LOD threshold is computed based on dataset size.

### Step 2 — Marker grouping

Markers are grouped into putative linkage groups using two-point recombination
frequency estimates (`onemap::rf_2pts()`). You set a **LOD threshold** and a
**maximum recombination frequency** (max RF); markers that meet both criteria are
placed in the same group.

The output is a table showing the number of markers assigned to each group and the
number of unlinked markers. You can adjust the LOD and max RF parameters and
re-run grouping until the result is biologically sensible (i.e. the number of groups
matches the expected haploid chromosome number of your organism).

### Step 3 — Marker ordering

Within each linkage group, markers are ordered to minimise the total map length.
Linkmapper offers three ordering algorithms:

- **RECORD** — Recombination Counting and Ordering. Recommended for most F2
  datasets; produces the most accurate order in simulation studies.
- **RCD** — Rapid Chain Delineation. Fast heuristic; suitable as a starting point
  or for very large groups.
- **UG (Unidirectional Growth)** — Alternative heuristic; useful when RECORD
  produces unexpected results.

Ordering is run on each linkage group in turn. The log-likelihood of the final
order is reported for each group, allowing you to compare algorithms.

### Step 4 — Linkage map generation

The ordered sequences are passed to `onemap::map()` to estimate inter-marker
distances using the selected mapping function (Kosambi or Haldane). The final
linkage map is displayed as:

- An **interactive plot** (plotly) where you can hover over markers to see their
  names and positions.
- A **static PNG** suitable for publication.

A summary table shows the number of markers and total length (cM) per linkage
group. Both the plot and the summary table are available for download.

### Step 5 — QTL analysis

If your data file contains phenotype columns, you can run a QTL scan on the
completed map. Choose a phenotype, a scanning method (interval mapping or CIM),
and a LOD significance threshold. The output is:

- A **LOD profile plot** showing the LOD score across the genome for the selected
  phenotype.
- A **QTL summary table** listing linkage group, position, and peak LOD score for
  all loci exceeding the threshold.

Both outputs are available for download.

---

## Downloading results

Results are available for download at each step:

| Step | Available downloads |
|---|---|
| Prior analysis | Missing data plot (PNG), segregation test table (CSV) |
| Marker grouping | Group assignment table (CSV) |
| Marker ordering | Ordered sequence summary (CSV) |
| Linkage map | Linkage map plot (PNG), marker statistics table (CSV) |
| QTL analysis | LOD profile plot (PNG), QTL summary table (CSV) |

---

## Limitations

- **Population types:** Only F2 intercross and backcross populations are currently
  supported. RILs, outcrossing populations, and polyploids are planned for future
  releases.
- **Session time limits:** The hosted ShinyApps.io version enforces free-tier
  session and compute limits. For large datasets or long-running analyses, install
  the R package and run locally.
- **Single-file input:** Linkmapper accepts one data file per session. Multi-file
  or multi-population analyses require running separate sessions.
- **QTL method:** The current QTL module uses the `qtl` package (interval mapping
  and CIM). Multi-trait or multi-QTL model selection is not yet implemented.

---

## Getting help

- **Bug reports and feature requests:** Open an issue on
  [GitHub](https://github.com/<!-- FILL: repo path -->/issues).
- **Usage questions:** See the tutorial vignette (`vignette("linkmapper-tutorial")`).
